---
title: "Lab 05 - La Quinta is Spanish for next to Denny's, Pt. 2"
author: "Hannah Crawley"
date: "2.11.25"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox) 
```

```{r load-data, message = FALSE}
states <- read_csv("data/states.csv")
```

### Exercise 1

> Filter the Denny’s data frame for Alaska (AK) and save the result as dn_ak. How many Denny’s locations are there in Alaska?

3 Dennys 


```{r}
dn_ak <- dennys %>% 
  filter(state == "AK")
nrow(dn_ak)
```
> Now, do the same for La Quinta data frame for Alaska (AK) and save the result as lq_ak. How many La Quinta locations are there in Alaska?

2 La Quintas

```{r}
lq_ak <- laquinta %>% 
  filter(state == "AK")
nrow(lq_ak)
```



### Exercise 2

> Next we will be calculating the distance between all Denny’s and all La Quinta locations in Alaska.

> How many pairings are there between all Denny’s and all La Quinta locations in Alaska, i.e., how many distances do we need to calculate between the locations of these establishments in Alaska? Calculate the number of pairings using the data frames you have already made

We will need to calculate 6 different distances (2x3)

### Exercise 3

> In order to calculate these distances, we need to first restructure our data to pair the Denny’s and La Quinta locations. To do so, we will join the two data frames. We have six join options in R. Each of these join functions take at least three arguments: x, y, and by.  
x and y are data frames to join
by is the variable(s) to join by

Four of these join functions combine variables from the two data frames:

Note: These functions are called mutating joins.

inner_join(): return all rows from x where there are matching values in y, and all columns from x and y.

left_join(): return all rows from x, and all columns from x and y. Rows in x with no match in y will have NA values in the new columns.

right_join(): return all rows from y, and all columns from x and y. Rows in y with no match in x will have NA values in the new columns.

full_join(): return all rows and all columns from both x and y, where there are not matching values, returns NA for the one missing.

```{r}
dn_lq_ak <- full_join(dn_ak, lq_ak, by = "state")
```
```{r}
dn_lq_ak
```


### Exercise 4

> How many observations are in the joined dn_lq_ak data frame? What are the names of the variables in this data frame.

Number of observations: 6 

Variables: address.x city.x, state, zip.x, longitude.x, latitude.x address.y city.y, zip.y, longitude.y, latitude.y

> .x in the variable names means the variable comes from the x data frame (the first argument in the full_join call, i.e. dn_ak), and .y means the variable comes from the y data frame. These variables are renamed to include .x and .y because the two data frames have the same variables and it’s not possible to have two variables in a data frame with the exact same name.

> Now that we have the data in the format we wanted, all that is left is to calculate the distances between the pairs.

### Exercise 5
 
> What function from the tidyverse do we use the add a new variable to a data frame while keeping the existing variables?


MUTATE! 

### Exercise 6

> Calculate the distances between all pairs of Denny’s and La Quinta locations and save this variable as distance. Make sure to save this variable in THE dn_lq_ak data frame, so that you can use it later.

```{r}
haversine <- function(long1, lat1, long2, lat2, round = 3) {
  # convert to radians
  long1 <- long1 * pi / 180
  lat1 <- lat1 * pi / 180
  long2 <- long2 * pi / 180
  lat2 <- lat2 * pi / 180

  R <- 6371 # Earth mean radius in km

  a <- sin((lat2 - lat1) / 2)^2 + cos(lat1) * cos(lat2) * sin((long2 - long1) / 2)^2
  d <- R * 2 * asin(sqrt(a))

  return(round(d, round)) # distance in km
}
```

```{r}
dn_lq_ak_distance <-  dn_lq_ak%>%
  mutate(
    distance = haversine(lat1 = latitude.x, long1 = longitude.x, lat2 = latitude.y, long2 = latitude.y)
  )
```

### Exercise 7

> Calculate the minimum distance between a Denny’s and La Quinta for each Denny’s location. To do so we group by Denny’s locations and calculate a new variable that stores the information for the minimum distance.

```{r}
dn_lq_ak_min <- dn_lq_ak_distance %>%
  group_by(address.x) %>%
  summarize(closest = min(distance))

```

### Exercise 8

> Describe the distribution of the distances Denny’s and the nearest La Quinta locations in Alaska. Also include an appropriate visualization and relevant summary statistics. 


```{r}
datasum_ak <- dn_lq_ak_distance %>% 
  summarise(
    mean_distance = mean(distance, na.rm = TRUE),
    max_distance = max(distance, na.rm = TRUE),
    count = n()
  )
print(datasum_ak)
```


### Exercise 9

> Repeat the same analysis for North Carolina: (i) filter Denny’s and La Quinta Data Frames for NC, (ii) join these data frames to get a complete list of all possible pairings, (iii) calculate the distances between all possible pairings of Denny’s and La Quinta in NC, (iv) find the minimum distance between each Denny’s and La Quinta location, (v) visualize and describe the distribution of these shortest distances using appropriate summary statistics.

```{r}
dn_nc <- dennys %>% 
  filter(state == "NC")
nrow(dn_nc)
```

```{r}
lq_nc <- laquinta %>% 
  filter(state == "NC")
nrow(lq_nc)
```
```{r}
dn_lq_nc <- full_join(dn_nc, lq_nc, by = "state")
```
```{r}
dn_lq_nc
```
```{r}
dn_lq_nc_distance <-  dn_lq_nc %>%
  mutate(
    distance = haversine(lat1 = latitude.x, long1 = longitude.x, lat2 = latitude.y, long2 = latitude.y)
  )
```


```{r}
dn_lq_nc_min <- dn_lq_nc_distance %>%
  group_by(address.x) %>%
  summarize(closest = min(distance))

```





### Exercise 10

> Repeat the same analysis for Texas.


### Exercise 11

> Repeat the same analysis for a state of your choosing, different than the ones we covered so far.



### Exercise 12

> Among the states you examined, where is Mitch Hedberg’s joke most likely to hold true? Explain your reasoning.



